---
title: "Actuarial Loss Prediction — Improvement Campaign 2026"
subtitle: "Four controlled rounds from the published model to the information ceiling"
author: "Eduardo Arias"
date: "2026-10-07"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

# Reproducibility statement

This report is **generated by the pipeline** (`targets::tar_make()` in the
project root): every table below is read from the pipeline's own targets,
which write `results/*.csv` as they run. Nothing here is transcribed by
hand. Environment: `rocker/tidyverse` container (R 4.6.1, Ubuntu jammy
binaries), seed 42 for the split and every model. The same pipeline
regenerates every number in this document with:

```{bash, eval=FALSE}
podman run --rm -v "$PWD":/home -w /home docker.io/rocker/tidyverse:latest \
  Rscript -e 'install.packages(c("targets","tarchetypes","tidymodels","tidytext",
  "SnowballC","stopwords","xgboost","glmnet","mgcv","rmarkdown")); targets::tar_make()'
```

The historical baseline (`analysis.html`, the 2025 course report) is a
separate document and is **not** regenerated: its numbers belong to its
original environment, as documented in the README's reproducibility note.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(targets)
r1 <- tar_read(round1)
r2 <- tar_read(round2)
r3 <- tar_read(round3)
r4 <- tar_read(round4)
fmt <- function(df, digits = 0) {
  df %>% mutate(across(where(is.numeric), ~ round(.x, digits)))
}
```

# Context

The published 2025 model (XGBoost, squared error on dollars, RMSE test
25,033.93 / MAE 7,257.61 — see `analysis.html`) was audited and then improved
in four controlled rounds, all on the same seeded 80/20 split (seed 42) of
the 54,000-claim Kaggle competition dataset, all metrics in dollars on the
same untouched test set.

# Round 1 — Clean protocol and objective (`experiments/01`)

```{r r1}
fmt(r1) %>% select(arm, eta, depth, rounds, RMSE, MAE, MAPE, R2, RMSLE) %>%
  knitr::kable(caption = "Round 1: audit-fixed protocol, three arms")
```

Finding: the gain came from the objective transform (log1p / Tweedie) and
text representation, **not** from fixing the leaky split — arm A (fixes
only) is worse than the faithful re-run of the original.

# Round 2 — Model families (`experiments/02`)

```{r r2}
fmt(r2) %>% select(arm, RMSE, MAE, MAPE, R2, RMSLE) %>%
  knitr::kable(caption = "Round 2: GLM/GAM/GAMM/hurdle vs XGBoost")
```

The Gamma GLM diverged (recorded as NA — documented failure). The hurdle
variant posts the best RMSE/R² of the round; the year random effect
contributes exactly nothing (GAM ≡ GAMM).

# Round 3 — Distribution study and target-encoded features (`experiments/03`)

```{r r3}
fmt(r3$metrics) %>% select(arm, RMSE, MAE, MAPE, R2, RMSLE) %>%
  knitr::kable(caption = "Round 3: TF-IDF vs target-encoded text severity")
```

Round-3 distribution findings (from the round's own study):
`InitialIncurredCalimsCost` explains **`r round(100 * r3$ceiling[["deciles"]], 1)`%**
(deciles) / **`r round(100 * r3$ceiling[["percentiles"]], 1)`%** (percentiles)
of the log-cost variance; the blow-up rate (>3×) is 24% in the cheapest
decile vs 6–7% in the top. Two target-encoded text columns replace 300
TF-IDF columns and beat them on every relative-error metric.

# Round 4 — Final squeeze (`experiments/04`)

```{r r4}
fmt(r4$metrics) %>% select(arm, RMSE, MAE, MAPE, R2, RMSLE, R2log) %>%
  knitr::kable(caption = "Round 4: OOF-TE, median objective, monotone, tuned, ensemble")
```

```{r r4-seg}
seg_df <- imap_dfr(r4$segments,
                   ~ tibble(arm = .y, segment = paste0("Q", seq_along(.x)),
                            mae = as.numeric(.x)))
seg_df %>%
  pivot_wider(names_from = arm, values_from = mae) %>%
  mutate(across(where(is.numeric), round)) %>%
  knitr::kable(caption = "Round 4: MAE by test-target quintile")
```

The median-objective model is the business pick (best MAE/MAPE and best
expensive-quintile MAE). Against the interaction-aware ceiling
(Initial × text-severity grouping, R²log ≈ 0.835) all arms land at
0.858–0.864: the remaining error is not in these features.

# Cumulative result

```{r cumulative}
pick <- function(metrics, arm) {
  row <- metrics %>% filter(arm == !!arm)
  if (nrow(row) == 0) return(tibble(RMSE = NA, MAE = NA, MAPE = NA))
  tibble(RMSE = round(row$RMSE[1]), MAE = round(row$MAE[1]), MAPE = round(row$MAPE[1], 1))
}
tribble(~model,
  "Published 2025 (original env.)",
  "R1 best (C_tweedie_tfidf)",
  "R2 best RMSE (HURDLE)",
  "R3 best (FE2_new_only)",
  "R4 business pick (A3_median)"
) %>%
  bind_cols(bind_rows(
    tibble(RMSE = 25033.93, MAE = 7257.61, MAPE = NA),
    pick(r1, "C_tweedie_tfidf"),
    pick(r2, "HURDLE"),
    pick(r3$metrics, "FE2_new_only"),
    pick(r4$metrics, "A3_median")
  )) %>%
  knitr::kable(caption = "Best model per round (test set, dollars)")
```

From the published original to the median-objective model: **MAE −27%,
MAPE −56%**, with the expensive-claims quintile improved ~7% — and the
remaining error characterized as structural (information ceiling), not a
modeling failure.
